Taking Dyads Seriously

### 2020-02-27 paper/1_intro.tex
r1 edits

\citet{aronow:etal:2015} estimate that during the period 2010 to 2015, over sixty articles were published in the \textit{American Political Science Review}, \textit{American Journal of Political Science}, and \textit{International Organization} using dyadic data.\footnote{In 2017, \textit{International Studies Quarterly} published a special issue on Dyadic Research Designs along with an online symposium to discuss the papers.} Most of these studies use a generalized linear model (GLM).  However, this approach to studying dyadic data increases the chance of faulty inferences by assuming data are conditionally independent and identically distributed (iid). Most standard approaches assume that the problems raised by non-iid relational data can be addressed by recalculating the standard errors of estimated parameters to reflect the potential clustering of cases. In practice, such strategies rarely work because they do not directly address the fundamental data generating process. This is important to consider since the inferential problems caused by non-iid  observations affect more than just diagonals of the variance co-variance matrix \citep{beck:2012,franzese:hays:2007,king:roberts:2014}.

In this article, we discuss and evaluate how well the Additive and Multiplicative Effects (AME) model can be used to account for the interdependencies underlying the data generating process of dyadic structures \citep{hoff:2008,minhas:etal:2018}. We focus on three types of interdependencies that can complicate dyadic analyses. First, dependencies may arise within a set of dyads if a particular actor is more likely to send or receive actions such as conflict.\footnote{In the case of undirected data where there is no clear sender or receiver, it is still essential to take into account the variance in how active actors are in the system.} Additionally, if the event of interest has a clear sender and receiver, we are likely to observe dependencies within a dyad; for example, if a rebel group initiates a conflict against a government, the government will likely reciprocate that behavior. We capture these effects, often referred to as first- and second-order dependencies, respectively, within the additive effects portion of the model. Third-order dependencies capture relationships of transitivity, balance, and clusterability between different dyads. For example, we can only understand why Poland was involved in a dyadic conflict with Iraq in 2003 if we understand that the United States invaded Iraq in 2003 and that Poland often participates in US-led coalitions. The multiplicative effects capture these sorts of dependencies, specifically, those that result because the specified model has not accounted for a latent set of shared attributes that affect actors' probability of interacting with one another.

>>>>

\citet{aronow:etal:2015} estimate that during the period 2010 to 2015, over sixty articles were published in the \textit{American Political Science Review}, \textit{American Journal of Political Science}, and \textit{International Organization} using dyadic data.\footnote{In 2017, \textit{International Studies Quarterly} published a special issue on Dyadic Research Designs along with an online symposium to discuss the papers.} Most of these studies use a generalized linear model (GLM).  However, this approach to studying dyadic data increases the chance of faulty inferences by assuming data are conditionally independent and identically distributed (iid). Most standard approaches assume that the problems raised by non-iid relational data can be addressed by recalculating the standard errors of estimated parameters to reflect the potential clustering of cases. \textcolor{blue}{In practice, such strategies remain insufficient because they do not directly model all relevant dependencies in the data\footnote{As \citet[p.15]{aronow:etal:2015} explain, “Of course, accounting for dynamic dyadic clustering may fail to fully account for all relevant dependencies in the data."} but instead adjust the variance estimator.\footnote{While this approach decreases the risk of a type 1 error, it increases risk of type 2 error (if the mean estimate is biased downwards, increasing the variance estimate will cause us to falsely accept the null)} The main goal of our study is to directly account for the relevant dependency structures in the data, and thus provide an accurate estimate while addressing both types of possible errors.} This is important to consider since the inferential problems caused by non-iid  observations affect more than just diagonals of the variance co-variance matrix \citep{beck:2012,franzese:hays:2007,king:roberts:2014}.

In this article, we discuss and evaluate how well the Additive and Multiplicative Effects (AME) model can be used to account for the interdependencies underlying the data generating process of dyadic structures \citep{hoff:2008,minhas:etal:2019}. We focus on three types of interdependencies that can complicate dyadic analyses. First, dependencies may arise within a set of dyads if a particular actor is more likely to send or receive actions such as conflict.\footnote{In the case of undirected data where there is no clear sender or receiver, it is still essential to take into account the variance in how active actors are in the system.} Additionally, if the event of interest has a clear sender and receiver, we are likely to observe dependencies within a dyad; for example, if a rebel group initiates a conflict against a government, the government will likely reciprocate that behavior. We capture these effects, often referred to as first- and second-order dependencies, respectively, within the additive effects portion of the model. Third-order dependencies capture relationships of transitivity, balance, and clusterability between different dyads. For example, we can only understand why Poland was involved in a dyadic conflict with Iraq in 2003 if we understand that the United States invaded Iraq in 2003 and that Poland often participates in US-led coalitions. The multiplicative effects capture these sorts of dependencies, specifically, those that result because the specified model has not accounted for a latent set of shared attributes that affect actors' probability of interacting with one another.
